---
title: "Introduction to Multilevel Measurement Models"
author: "SMIP Summer School 2025: Lecture 01" 
format: 
  revealjs:
    multiplex: true
    footer: "SMIP Summer School 2025, Multilevel Measurement Models, Lecture 01"
    theme: ["pp.scss"]
    slide-number: c/t
    incremental: false
editor: source
--- 

## About Your Instructors

- [Lesa Hoffman](https://lesahoffman.com){target="_blank"}: lesa-hoffman@uiowa.edu
- [Jonathan Templin](https://jonathantemplin.com){target="_blank"}: jonathan-templin@uiowa.edu

## Workshop Introduction
This is an ambitious workshop:

* Measurement models (Item Response Theory)
* Multilevel models for clustered data
* Bayesian statistics
* Multilevel measurement models

Each could be its own workshop

## Workshop Materials

Set of slides on topics pulled from other course materials

* In case additional detail on any topic is needed

Full set of analysis files in R and Stan

* In case you would like to work ahead

## Workshop Schedule

* Mornings: 9:00-12:30 (**Including Friday**)
  * Break at 10:45
* Afternoons:
  * Tuesday: 14:00-17:15 (break at 16:00)
  * Wednesday: 15:15-17:45 (break at 16:45)
  * Thursday: 14:00-15:45 (no break)

## Lecture and Example Time Plan

Last hour of each day is open

Your choice:

* Questions
* Example practice
* Work with your data
* Leave early

## Syntax and Model Note

* As stan models can take a very long time to run, all prior model results have been saved
* All lectures are in self-contained folders 
* All files are available at [https://github.com/jonathantemplin/MultilevelMeasurementModels2025](https://github.com/jonathantemplin/MultilevelMeasurementModels2025){target="_blank"}

## Workshop Sections

1. Introduction to Measurement Models, Bayesian Statistics, and Stan
2. Introduction to Multilevel Models
3. Introduction to Multilevel Measurement Models

# Running Example

## Running Example Data

Data come from a secondary school (10th grade) end-of-grade mathematics assessment given around the year 2006 in a midwestern state in the USA

* Sample of 50 students from 62 schools
* Sample of 10 mathematics items from the assessment
  * Items are scored correct/incorrect (```score1```-```score10```)
  * The sum of all 10 items is included (```sumScore```)
* Other variables:
  * ```studentID```: Student ID number (created for this example)
  * ```schoolID```: School ID number (created for this example)
  * ```frlunch```: Free/reduced lunch status (1 = free/reduced lunch, 0 = not free/reduced lunch; indirect indicator of student socioeconomic status)


## Distribution of Sum Scores

```{r sumScore, echo=FALSE, fig.height=4, fig.width=6}
load("modelingData.RData")
hist(modelingData$sumScore)
```

## Distribution of Item Difficulties

```{r itemScores, echo=FALSE, fig.height=4, fig.width=6}
hist(apply(X=modelingData[paste0("score", 1:10)], MARGIN=2, FUN=mean), xlab = "Item Difficulty", main = "")

```

## Workshop Lectures

- Lecture 01: Introduction to Multilevel Measurement Models Workshop
- Lecture 02: Introduction to Measurement Models
- Lecture 03: Introduction to Bayesian Statistics
- Lecture 04: Introduction to MCMC and Stan
- Lecture 05: Bayesian IRT Models in Stan
- Lecture 06: Introduction to Multilevel Models
- Lecture 07: Bayesian Multilevel Models in Stan
- Lecture 08: Bayesian Multilevel Measurement Models in Stan


